\documentclass[11pt]{letter}
\usepackage{geometry}
\usepackage{times}

\begin{document}

\thispagestyle{empty}

\begin{flushleft}
   Editor-in-Chiefs\\
   Transportation Research Part E\\
   \vspace{1cm}
   \today
\end{flushleft}


Dear Professor Tsan-Ming Choi and Professor Shuaian Wang,

We have revised and are resubmitting our previous rejected submission, TRE-D-23-01128, for your re-consideration.  Our changes to the manuscript in response to the comments are highlighted below.

1. Comment from Editor: \textit{As a routine check, your paper is scanned for similarity. The similarity index is 26\% (report attached or available online). Given the high similarity, your paper is returned.}

We have updated the manuscript to significantly reduce the similarity index.  Excluding references, the similarity index of our manuscript is 10\% according to iThenticate.  Excluding Table 1, which includes authors' naming of MTVRPTW variants, the similarity index is 6\%.  Furthermore, none of the similarity scores with a single source is above 1\%.  The similarity score of 1\% is due to common phrases and names of the methods in the literature.

2. Comment from Editor: \textit{The review doesn't seem to bring a major contribution to logistics. Whether this review relates to TRE is also under doubt as most reviewed papers are from other journals. It is important to clearly position this paper's position in logistics and explains why it belongs to TRE. Our associate editor can decline papers simply because of relevance.}

We reviewed 5 additional papers from TRE that are directly relevant to the problem (MTVRPTW). We first updated the Introduction with further applications of MTVRPTW to show its application in logistics clearly. We further reviewed these papers in the classifications of MTVRPTW, exact algorithms, and heuristic algorithms, where applicable.

Here is the list of references for the new papers from TRE that we reviewed:

\begin{itemize}
\item Christiansen, M., Fagerholt, K., Rachaniotis, N. P., and St{\aa}lhane, M. (2017). Operational planning of routes and schedules for a fleet of fuel supply vessels. Transportation Research Part E: Logistics and Transportation Review, 105, 163–175.

\item Huang, N., Li, J., Zhu, W., and Qin, H. (2021). The multi-trip vehicle routing problem with time windows and unloading queue at depot. Transportation Research Part E: Logistics and Transportation Review, 152, 102370

\item Liu, S., Qin, S., and Zhang, R. (2018). A branch-and-price algorithm for the multi-trip multi-repairman problem with time windows. Transportation Research Part E: Logistics and Transportation Review, 116, 25–41.

\item Tang, J., Yu, Y., and Li, J. (2015). An exact algorithm for the multi-trip vehicle routing and scheduling problem of pickup and delivery of customers to the airport. Transportation Research Part E: Logistics and Transportation Review, 73, 114–132.

\item Zhen, L., Ma, C., Wang, K., Xiao, L., and Zhang, W. (2020). Multi-depot multi-trip vehicle routing problem with time windows and release dates. Transportation Research Part E: Logistics and Transportation Review, 135, 101866.

\end{itemize}
\vspace{1cm}

Sincerely,

Thanh Trong Truong Nguyen\\
National University of Singapore\\
truongthanh@u.nus.edu
%Tsung-Sheng Chang\\
%Professor\\
%National Yang Ming Chiao Tung University\\
%tsc@nycu.edu.tw


\end{document}
